The impact of the COVID-19 pandemic on academic productivity

### 2022-06-21 article/ms.tex
add zenodo link, final comments

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics are posting more papers on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in quantitative biology where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians). 
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We retrieved metadata for 1,660,423 pre-prints posted to \arxiv\ between 1 April 2007 and 31 May 2022. The metadata includes the creation date, research field(s), title, author name(s), abstract, and other miscellaneous information\cite{Clement:2019}. There is an increasing number of pre-prints posted to \arxiv\ each year in nearly every field (Figure~\ref{fig:1}). These long-term trends are relatively predictable from year to year, allowing us to quantify any change in academic productivity due to the COVID-19 pandemic. \add{The number of pre-prints posted varies substantially by category. Computer science (\texttt{cs}) is the growing the fastest with more than 5,000 pre-prints per month, followed by math (\texttt{math}) with about 3,500 per month. For contrast, only $\approx$200 pre-prints are posted each month quantitative biology (\texttt{q-bio}), and just $\approx$40 in nuclear experimental physics (\texttt{nucl-ex}). For this reason, we model the number of monthly pre-prints in each research field separately.} We used the number of publications from January 2015 to December 2019 to train a predictive Gaussian process model\cite{Rasmussen:2006}. This model treats the number of monthly pre-prints in a research field as a sequence of normally distributed random variables. The expected number of pre-prints was assumed to increase linearly with time, and a quasi-periodic kernel function\cite{Rasmussen:2006,Ambikasaran:2014} was used to capture seasonal variations (Figure~\ref{fig:1}). In nearly all research fields, the number of pre-prints posted since January 2020 agree closely with the model predictions (Table~1) indicating no significant overall decline in academic productivity due to the COVID-19 pandemic. %In fact, the total number of \arxiv\ pre-prints in 2020 across all fields (168,630) exceeds our model predictions ($162{,}577 \pm 4{,}393$; $+1.4\sigma$).  % todo

\add{The exception is physics (\texttt{physics}), which shows the largest decline in pre-prints by number since the pandemic began. In 2020 the number posted to physics actually exceeded our model predictions (15,242 compared to 14,746) by $1\sigma$, driven by a short peak of pre-prints posted to Physics and Society (\texttt{physics.soc-ph}) in early 2020. When we segment the number of pre-prints posted to physics by their sub-category we find that Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}) had significant drops in pre-print counts since the pandemic began. All other physics sub-categories remained largely stable with time. The category of mathetmatical physics (\texttt{math-ph}) also showed a decline in pre-prints since the pandemic began, but this category has about an order of magnitude fewer pre-prints posted than physics. The mathematical physics category does not have defined sub-categories, making it more difficult to evaluate whether this decline occurred in particular areas of study}.

Experimental research projects often require specialised laboratories, or data to be collected over many years. The pandemic has forced many laboratories to close or operate with restricted access\cite{PhysicsWorld,ScienceMag,Castelvecchi}, which could lead to long-lasting delays in ongoing experiments or immediate drops in publication rates in part due to difficulty in accessing completed (or nearly completed) experiment data. \add{This could explain why there are declines in pre-prints in some sub-categories (e.g., Applied Physics), but there is not significant evidence of this in pre-print numbers of experimental-only fields (e.g., \texttt{hep-ex} or \texttt{nucl-ex}). In those fields the model predictions agree reasonably well with the pre-prints since 2020.}

@@

\begin{figure}[!t]
\includegraphics[width=0.95\linewidth]{physics-pre-prints}
\caption{\add{The number of pre-prints posted to \arxiv\ in the categories of Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}). The decline in physics pre-prints since the pandemic began (Figure~\ref{fig:1}) is largely explained by the drop in pre-prints in Computational Physics and Applied Physics.}}
\label{fig:physics}
\end{figure}

@@

The COVID-19 pandemic appears to have had minimal impact on collective research productivity to date.
\add{However, some significant changes were noted within specific areas of research. The largest persistent decline in pre-prints has occurred in (primarily computational and applied) physics. As of mid-2022, this steady decline continues.} Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020. In this study we have only focussed on the quantity of publications as a measure for productivity, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

\dataccess{\add{All of the data and software used in this work are publicly available. The dataset used here is from \arxiv. The original dataset is curated by Cornell University and hosted by Kaggle (https://www.kaggle.com/Cornell-University/arxiv). The dataset is updated weekly. We accessed this dataset on 06 June 2022. We have stored a time-stamped version of this dataset, as well as the code used to process the data, as Zenodo record \cite{ZenodoRecord}.
A public GitHub repository (https://github.com/andycasey/arxiv) stores all software used to preprocess and analyse these data, including the scripts to produce the figures shown here, and to compile this manuscript. The repository includes a \texttt{README.md} file that describes the software environment used to perform the analysis, and steps on how to reproduce our work in full.}}

\aucontribute{\add{AC and IM conceived the idea to examine changes in academic productivity due to the pandemic. PK and IM helped guide the investigation and provided ideas on new avenues to explore and tests to perform. AC conducted the analysis. AC led the writing of the manuscript, with all authors contributing significantly.} All authors read and iterated upon the text of the final manuscript.}

@@

\bibitem{Viglione:2020}
{Viglione, G.}, {Are women publishing less during the pandemic? Here?s what the data say}. \emph{Nature}, \textbf{581}, 356-366, doi:10.1038/d41586-020-01294-9 \newblock (2020).

@@

>>>>

%%%% Abstract text to be placed here %%%%%%%%%%%%
\begin{abstract}
`Publish or perish’ is an expression describing the pressure on academics to consistently publish research to ensure a successful career in academia. 
With a global pandemic that has changed the world, how has it changed academic productivity? 
Here we show that academics are posting more papers on the arXiv pre-print server than would be expected had the pandemic not occurred: 341,474 pre-prints were posted between 2020-2021, up 14\% from 2019 and above the predictions of our Gaussian process model.
%168,630 were posted in 2020 , a +12.6\% change from 2019 and $+1.4\sigma$ deviation above the 162,577 predicted by our Gaussian process model. % todo: reword
We find that that most individual fields were not measurably affected by the pandemic.
However, we do observe a significant ongoing decline in \add{general} physics pre-prints, as well as temporary declines in areas of high-energy physics, where the cancellation of a key conference played a key role.
The most significant increase occurred in quantitative biology where a 50\%~increase ($+8\sigma$) in pre-prints posted in 2020 was driven by research related to the COVID-19 pandemic. 
Some of these `COVID papers' are by biologists using arXiv for the first time, but many are written by researchers from other fields (e.g., physicists, mathematicians). 
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

We retrieved metadata for 1,660,423 pre-prints posted to \arxiv\ between 1 April 2007 and 31 May 2022. The metadata includes the creation date, research field(s), title, author name(s), abstract, and other miscellaneous information\cite{Clement:2019}. There is an increasing number of pre-prints posted to \arxiv\ each year in nearly every field (Figure~\ref{fig:1}). These long-term trends are relatively predictable from year to year, allowing us to quantify any change in academic productivity due to the COVID-19 pandemic. The number of pre-prints posted varies substantially by category. Computer science (\texttt{cs}) is the growing the fastest with more than 5,000 pre-prints per month, followed by math (\texttt{math}) with about 3,500 per month. For contrast, only $\approx$200 pre-prints are posted each month quantitative biology (\texttt{q-bio}), and just $\approx$40 in nuclear experimental physics (\texttt{nucl-ex}). For this reason, we model the number of monthly pre-prints in each research field separately. We used the number of publications from January 2015 to December 2019 to train a predictive Gaussian process model\cite{Rasmussen:2006}. This model treats the number of monthly pre-prints in a research field as a sequence of normally distributed random variables. The expected number of pre-prints was assumed to increase linearly with time, and a quasi-periodic kernel function\cite{Rasmussen:2006,Ambikasaran:2014} was used to capture seasonal variations (Figure~\ref{fig:1}). In nearly all research fields, the number of pre-prints posted since January 2020 agree closely with the model predictions (Table~1) indicating no significant overall decline in academic productivity due to the COVID-19 pandemic. %In fact, the total number of \arxiv\ pre-prints in 2020 across all fields (168,630) exceeds our model predictions ($162{,}577 \pm 4{,}393$; $+1.4\sigma$).  % todo

The exception is \add{general physics (\arxiv\ category \texttt{physics}; not all physics-related domains}), which shows the largest decline in pre-prints by number since the pandemic began. In 2020 the number posted to general physics actually exceeded our model predictions (15,242 compared to 14,746) by $1\sigma$, driven by a short peak of pre-prints posted to Physics and Society (\texttt{physics.soc-ph}) in early 2020. When we segment the number of pre-prints posted to physics by their sub-category we find that Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}) had significant drops in pre-print counts since the pandemic began (Figure~\ref{fig:physics}). All other physics sub-categories remained largely stable with time. The category of mathetmatical physics (\texttt{math-ph}) also showed a decline in pre-prints since the pandemic began, but this category has about an order of magnitude fewer pre-prints posted than physics. The mathematical physics category does not have defined sub-categories, making it more difficult to evaluate whether this decline occurred in particular areas of study.

Experimental research projects often require specialised laboratories, or data to be collected over many years. The pandemic has forced many laboratories to close or operate with restricted access\cite{PhysicsWorld,ScienceMag,Castelvecchi}, which could lead to long-lasting delays in ongoing experiments or immediate drops in publication rates in part due to difficulty in accessing completed (or nearly completed) experiment data. This could explain why there are declines in pre-prints in some sub-categories (e.g., Applied Physics), but there is not significant evidence of this in pre-print numbers of experimental-only fields (e.g., \texttt{hep-ex} or \texttt{nucl-ex}). In those fields the model predictions agree reasonably well with the pre-prints since 2020.

@@

\begin{figure}[!t]
\includegraphics[width=0.95\linewidth]{physics-pre-prints}
\caption{The number of pre-prints posted to \arxiv\ in the categories of Computational Physics (\texttt{physics.comp-ph}) and Applied Physics (\texttt{physics.app-ph}). The decline in physics pre-prints since the pandemic began (Figure~\ref{fig:1}) is largely explained by the drop in pre-prints in Computational Physics and Applied Physics.}
\label{fig:physics}
\end{figure}

@@

The COVID-19 pandemic appears to have had minimal impact on collective research productivity to date.
However, some significant changes were noted within specific areas of research. The largest persistent decline in pre-prints has occurred in \add{general physics (arXiv primary category \texttt{physics}), largely due to a decline in  pre-prints posted to computational and applied physics areas within the \texttt{physics} primary category. As of mid-2022, this steady decline continues.} Amongst individual fields the largest \emph{relative} decline in a single field is explainable by a high energy lattice physics conference that was cancelled in 2020. In this study we have only focussed on the quantity of publications as a measure for productivity, and not their quality. While citations are the most used measure of impact of academic publications, that metric becomes a more biased statistic when many related pre-prints are all being posted nearly at the same time\cite{Fassin:2021}. The relatively long timescales of academic research would suggest that the full impact on academic productivity due to the COVID-19 pandemic is yet to be seen. The methodology and results presented here provide a quantitative framework for future investigations of the longer term impacts of the pandemic.

@@

\dataccess{All of the data and software used in this work are publicly available. The dataset used here is from \arxiv. The original dataset is curated by Cornell University and hosted by Kaggle (https://www.kaggle.com/Cornell-University/arxiv). The dataset is updated weekly. We accessed this dataset on 06 June 2022. We have stored a time-stamped version of this dataset, as well as the code used to process the data \cite{ZenodoRecord}.
A public GitHub repository (https://github.com/andycasey/arxiv) stores all software used to preprocess and analyse these data, including the scripts to produce the figures shown here, and to compile this manuscript. The repository includes a \texttt{README.md} file that describes the software environment used to perform the analysis, and steps on how to reproduce our work in full.}

\aucontribute{AC and IM conceived the idea to examine changes in academic productivity due to the pandemic. \add{PR} and IM helped guide the investigation and provided ideas on new avenues to explore and tests to perform. AC conducted the analysis. AC led the writing of the manuscript, with all authors contributing significantly. All authors read and iterated upon the text of the final manuscript.}

@@

\bibitem{Viglione:2020}
{Viglione, G.}, {Are women publishing less during the pandemic? Here\add{'}s what the data say}. \emph{Nature}, \textbf{581}, 356-366, doi:10.1038/d41586-020-01294-9 \newblock (2020).

@@

\add{\bibitem{ZenodoRecord} Casey, Andrew R. (2022). Impact of the COVID-19 pandemic on academic productivity [Data set]. Zenodo. https://doi.org/10.5281/zenodo.6615684}
